Staging Abstract Interpreters

### 2019-07-07 paper/10_intro.tex
paper

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level programming approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses the interpretive
overhead, e.g., pattern matching on the ASTs and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program once, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries. Second, the
abstract interpreter written in pure languages usually extensively uses effect
systems to model the behaviors of abstract interpretation, such as
nondeterminism. For example, \citet{DBLP:journals/pacmpl/DaraisLNH17} and
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract interpreters use
monad transformers; the very recent \citet{Githubsemantic}'s \textsc{Semantic} framework
uses extensible effects. With certain merits and elegance of the implementation,
the effect systems may also introduce additional performance overhead.

@@

Our proposed framework adopts the type-based multi-stage programming
and implements the first Futamura projection of a big-step abstract
interpreter for a stateful, higher-order core language.  We use
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s abstract definitional
interpreter that uses monad transformers as the unstaged baseline. By
deriving staged monads that can be used to do code generation, we may
obtain a staged abstract interpreter that fulfills the specialization.
After staging, the generated code is specialized to the input program,
and all monadic operations are all inlined.

>>>>

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties with these high-level programming approaches.
First, similar to concrete interpreters, the abstract interpreter analyzes the
program by traversing the abstract syntax tree, which poses the interpretive
overhead, e.g., pattern matching on the ASTs and recursive calls on the sub
expressions. Those kind of overhead can be negligible if the abstract
interpreter only runs on the program once, but also can be accumulated
significantly if it runs repeatedly, for example, on libraries. Second, the
abstract interpreter written in pure languages usually extensively uses effect
systems to model the behaviors of abstract interpretation, such as
nondeterminism. For example, \citet{DBLP:journals/pacmpl/DaraisLNH17} and
\citet{Sergey:2013:MAI:2491956.2491979}'s monadic abstract interpreters;
the very recent \citet{Githubsemantic}'s \textsc{Semantic} framework
uses extensible effects. With certain merits and elegance of the implementation,
the effect systems may also introduce additional performance overhead.

@@

Our proposed framework adopts the type-based multi-stage programming
and implements the first Futamura projection of a big-step abstract
interpreter for a stateful, higher-order core language.  We use
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s abstract definitional
interpreter that uses monad transformers as the unstaged baseline. By
deriving staged monads that can be used to do code generation, we may
obtain a staged abstract interpreter that fulfills the specialization.
After staging, the generated code is specialized to the input program,
and all monadic operations are all inlined and compiled down to low-level
Scala code.

### 2019-07-07 paper/70_eval.tex
paper

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\ 
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening (sec.)}  &  \multicolumn{3}{c}{w/ store-widening (ms)}\\
    \midrule
    @fib@               & 32   & 0.013      & 0.002      & 6.50x      & 1.537 & 0.067  &  22.87x       \\
    @rsa@               & 451  & 0.232      & 0.042      & 5.52x      & 15.113& 1.449 &  10.43x       \\
    @church@            & 120  & 67.342     & 18.917     & 3.56x      & 2.962 & 0.291 &  10.19x       \\
    @fermat@            & 310  & 0.024      & 0.004      & 6.00x      & 7.624 & 0.622 &  12.26x       \\
    @mbrotZ@            & 331  & 0.763      & 0.127      & 6.01x      & 12.229& 0.729 &  16.77x       \\
    @lattice@           & 609  & 31.874     & 8.816      & 3.62x      & 17.943& 0.909 &  19.75x      \\
    @kcfa-worst-16@     & 182  & 0.045      & 0.006      & 7.50x      & 4.658 & 0.785 &  5.93x     \\
    @kcfa-worst-32@     & 358  & 0.284      & 0.022      & 12.91x     & 10.005& 1.680 &  5.95x     \\
    @kcfa-worst-64@     & 774  & 2.18       & 0.129      & 16.90x     & 22.999& 3.872 &  5.94x     \\
    @solovay-strassen@  & 523  & 5.094      & 1.189      & 4.28x      & 19.118& 0.654 &  29.25x      \\
    @regex@             & 550  & -          & -          & -          & 7.133 & 0.624 &  11.43x      \\
    @matrix@            & 1891 & -          & -          & -          & 86.176& 4.592 &  18.76x      \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on two monovariant CFA (without and with store-widening). The
\textit{unstaged} and \textit{staged} columns show the median time to finish the
analysis. The column $\frac{\text{unstaged}}{\text{staged}}$
shows the improvement from staging. The dash - in some cells indicates timeout.
As we can see from the table, the staged version significantly outperforms the
unstaged version on all benchmarks. %, especially for the @kcfa-worst-@$n$ series.
We observe the most improvement of 29x times faster than the unstaged version on
@solovay-strassen@. On average, the staging produces code that runs 11x times faster
than the unstaged analyzer.

@@

\iffalse
As we can see from the table, the staged version significantly outperforms the
unstaged especially for the difficult ones. However, for some benchmarks, we
observe degraded performance. We conjecture this happens due to the fact that
the large size of the generated code is unoptimized -- as what we test here is a
simple implementation almost identical to what we described in the previous
sections, we believe the performance still can be improved if more engineering
effort is taken, such as optimizing code generation and avoiding code
duplication.
\fi

>>>>

\begin{figure}[h]
\footnotesize
\begin{tabular}{@{}ll|lll|lll@{}}
\toprule
    program             &\#AST & unstaged   & staged     & $\frac{\text{unstaged}}{\text{staged}}$ & unstaged   & staged    & $\frac{\text{unstaged}}{\text{staged}}$  \\ 
    \midrule
                        &      & \multicolumn{3}{c}{w/o store-widening (sec.)}  &  \multicolumn{3}{c}{w/ store-widening (ms)}\\
    \midrule
    @fib@               & 32   & 0.013      & 0.002      & 6.50x      & 1.537 & 0.067  &  22.87x      \\
    @rsa@               & 451  & 0.232      & 0.042      & 5.52x      & 15.113& 1.449 &  10.43x       \\
    @church@            & 120  & 67.342     & 18.917     & 3.56x      & 2.962 & 0.291 &  10.19x       \\
    @fermat@            & 310  & 0.024      & 0.004      & 6.00x      & 7.624 & 0.622 &  12.26x       \\
    @mbrotZ@            & 331  & 0.763      & 0.127      & 6.01x      & 12.229& 0.729 &  16.77x       \\
    @lattice@           & 609  & 31.874     & 8.816      & 3.62x      & 17.943& 0.909 &  19.75x       \\
    @kcfa-worst-16@     & 182  & 0.045      & 0.006      & 7.50x      & 4.658 & 0.785 &  5.93x        \\
    @kcfa-worst-32@     & 358  & 0.284      & 0.022      & 12.91x     & 10.005& 1.680 &  5.95x        \\
    @kcfa-worst-64@     & 710  & 2.18       & 0.129      & 16.90x     & 22.999& 3.872 &  5.94x        \\
    @solovay-strassen@  & 523  & 5.094      & 1.189      & 4.28x      & 19.118& 0.654 &  29.25x       \\
    @regex@             & 550  & -          & -          & -          & 7.133 & 0.624 &  11.43x       \\
    @matrix@            & 1732 & -          & -          & -          & 86.176& 4.592 &  18.76x       \\
    \bottomrule
\end{tabular}
\caption{Evaluation result for monovariant control-flow analysis.} \label{evaluation_result}
\end{figure}

Figure~\ref{evaluation_result} shows the evaluation result, comparing the
performance improved by staging on two monovariant CFA (without and with store-widening). The
\textit{unstaged} and \textit{staged} columns show the median time to finish the
analysis. The column $\frac{\text{unstaged}}{\text{staged}}$
shows the improvement from staging. The dash - in some cells indicates timeout.
As we can see from the table, the staged version significantly outperforms the
unstaged version on all benchmarks.
We observe the most improvement of 29x times faster than the unstaged version on
@solovay-strassen@. On average, the staging produces code that runs 11x times faster
than the unstaged analyzer.

@@
